---
title: "Business Insights Report"
subtitle: "Data Visualization and Communication"
author: "Victor--6677706903"
date: "`28 OCT 2025`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---
```{r}
getwd()
```
### Question 1

```{r}
library(ggplot2)
business_data <- read.csv("business_performance.csv")
```

##A
```{r}
str(business_data)
summary(business_data)
```

The data in quarter and channel are categorical factor variables 

The range is 412125-150231= 261894A

##B

```{R}
Total_revenue_by_channel <- aggregate(revenue ~ channel,data = business_data,sum)

Avg_customers_by_quarter <- aggregate(customers ~ quarter,data = business_data,mean)

Total_revenue_by_channel
Avg_customers_by_quarter
```

The Online Channel generates the most revenue 
Q3 has the most customers on avaerage 
This suggest that this period of the year attracts the most customer traffic in comparision to the other quarters 

##C

```{r}
business_data$revenue_per_customer <- business_data$revenue / business_data$customers

avg_rev_per_customer <- aggregate(revenue_per_customer~channel,data = business_data, mean)

avg_rev_per_customer$revenue_per_customer <- round(avg_rev_per_customer$revenue_per_customer, 2)

avg_rev_per_customer
```

Retail channel has the highest average revenue per customer 
This informs us that on avarage an individual customer will spend more in store than online.

###Question 2

##A

```{R}
channel_totals <- aggregate(revenue ~ channel, data = business_data, sum)

ggplot(channel_totals, aes(x = channel, y = revenue)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Revenue by Channel", 
       x = "Sales Channel", y = "Revenue ($)")
```

The Online channel should receive more investment because it generates the highest revenue compared to both the Mobile and Retail channels 

##B

```{r}
ggplot(business_data, aes(x = customers, y = revenue, color = channel)) +
  geom_point(size = 3) +
  labs(title = "Customer Count vs Revenue by Channel",
       x = "Number of Customers", y = "Revenue ($)")
```

The Retail Channel seems to be the most efficient at converting customers into revenue because though they are a few they are all close together suggesting its the most efficient 


###Question 3 

##A

```{R}
library(knitr)


performance_summary <- aggregate(
  cbind(revenue, customers, conversion_rate) ~ channel,
  data = business_data,
  FUN = mean
)


total_revenue <- aggregate(revenue ~ channel, data = business_data, sum)
performance_summary$Total_Revenue <- total_revenue$revenue


performance_summary$Revenue_per_Customer <- round(
  performance_summary$revenue / performance_summary$customers, 2
)


performance_summary$revenue <- round(performance_summary$revenue, 0)
performance_summary$customers <- round(performance_summary$customers, 0)
performance_summary$conversion_rate <- round(performance_summary$conversion_rate, 2)
performance_summary$Total_Revenue <- round(performance_summary$Total_Revenue, 0)

kable(
  performance_summary,
  col.names = c("Sales Channel",
                "Avg Revenue ($)",
                "Avg Customers",
                "Avg Conversion Rate (%)",
                "Total Revenue ($)",
                "Revenue/Customer ($)"),
  digits = 2,
  caption = "Sales Channel Performance Summary"
)
```

Online has the highest average revenue $347961 and revenue per customer $1,567 In-store (Retail) has the lowest revenue per customer $543 So I recommend Increasing digital marketing and personalized promotions for in-store customers to boost conversion efficiency. Although the In-Store channel converts 5% of customers, its revenue per customer ($543) lags behind Online’s $1,567. Replicating Online’s upselling and personalization strategies in stores could raise average spend per customer, potentially increasing total revenue by over 15% and improving overall business profitability.


---
title: "Government Data Analysis: [Business insight report]"
---

[Your existing analysis content here]

---

[← Back to Projects](projects.html) | [← Back to Home](index.html)